Canopy leaf area of an evergreen Eucalyptus woodland does not respond to elevated

### 2015-05-04 manuscript.Rmd
Manuscript format OK; captions all done; proper sub and superscripts; template uses pagebreak before Header1; Modified FigureSI2

---
title: Canopy leaf area of an evergreen Eucalyptus woodland does not respond to elevated
  atmospheric [$CO_2$] but tracks water availability
author: ""
date: ""
output:
  word_document:
    reference_docx: manuscriptstyle.docx
  pdf_document: default
  html_document:
    number_sections: yes
csl: global-change-biology.csl
bibliography: references.bib
---

@@

library(broom)
library(lme4)
library(lmerTest)
library(car)

@@

# Some definitions of constants
.gambase <- 18

@@

@@

@@

\
$\mathrm{Remko~A.~Duursma}^1$, $\mathrm{Teresa~E.~Gimeno}^{1,2}$, $\mathrm{Matthias~M.~Boer}^1$, $\mathrm{Kristine~Y.~Crous}^1$, $\mathrm{Mark~G.~Tjoelker}^1$, $\mathrm{David~S.~Ellsworth}^1$

\
$~^1$ Hawkesbury Institute for the Environment, University of Western Sydney, Locked Bag 1797, Penrith, NSW, Australia

\
$~^2$ INRA, UMR 1391 ISPA, 33140 Villenave d'Ornon, France

@@

Leaf area, quantified by the leaf area index (*L*), is a crucial driver of forest productivity, water use, and energy balance. Because *L* responds to environmental drivers, it can represent an important feedback to climate change, but its responses to rising [$CO_2$] and water availability of unmanaged forests have been poorly quantified. We studied canopy leaf area dynamics for `r nummon` months in a native evergreen *Eucalyptus* woodland exposed to Free Air $CO_2$ Enrichment (the EucFACE experiment), in a temperate (hot summer) climate where water limitation is common. We hypothesized that, because of expected higher productivity and increased water-use efficiency, *L* should increase with elevated [$CO_2$]. We estimated *L* from diffuse canopy transmittance, and measured monthly leaf litter production. Contrary to expectation, *L* did not respond to elevated [$CO_2$] in this timeframe. We found that *L* varied between `r rangeLAI[1]` and `r rangeLAI[2]`, but did not show a clear annual pattern. Instead, dynamics of *L* showed a quick increase after heavy rainfall, and a steady during periods of low rainfall. Leaf litter production was correlated to changes in *L*, both during periods of decreasing *L* (when no leaf growth occurred), and during periods of increasing *L* (active shedding of old foliage when new leaf growth occurred). Leaf lifespan, estimated from mean *L* and total annual litter production, was significantly increased under elevated [$CO_2$] (p = 0.051). We discuss these results in the context of modelling phenology of evergreen woodlands in response to environmental drivers.

@@

Leaf area index (*L*) is an important ecosystem attribute underlying productivity of vegetation, [@Reich2012], and a key variable in land surface models [@Kala2013; @Walker2014]. *L* is a dynamic canopy property, as it is generally responsive to variation in water and nutrient availability, as demonstrated in experiments [@Herbert1995; @Hubbard2004; @White2010] and observational studies across rainfall gradients [@Ellis2008; @Iio2013; @Jin2013]. Ecosystem models generally predict that *L* increases in response to elevated atmospheric [$CO_2$] ($eC_a$) due to increased net carbon uptake [@DeKauwe2014; @Hatton1992; @Kergoat2002; @Zhu2012]. Also expected because increased quantum yield of photosynthesis in $eC_a$ would allow leaf retention into deeper shade [@Norby1999], but little empirical evidence is available on the response of *L* to $eC_a$ in mature forests [@Norby2011]. Quantifying the magnitude and direction of the $eC_a$ effect on *L* is crucial, as changes in *L* associated with climate change drivers such as increases in $C_a$ may provide an important feedback to climate change, and modify direct physiological responses to $C_a$ [@Tor-ngern2015; @Murray2012].

\
Direct effects of $eC_a$ on vegetation processes include an enhancement of photosynthesis and productivity, a decrease in stomatal conductance and leaf-level water use and an increase in the water-use efficiency [@Ainsworth2005; @Ellsworth2012]. Although numerous studies have been conducted on the response of physiology and plant growth to $eC_a$, very few of these inform the response of steady-state *L* to $eC_a$, because it requires a sufficiently mature forest to be exposed to free air $CO_2$ enrichment (FACE) [@Norby2011]. Increased carbon availability through enhanced photosynthesis under $eC_a$ does not necessarily lead to increases in *L*. Out of the six forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch) and needle-leaved evergreen pine stands showed increased *L* in response to $eC_a$ *L* [@Norby2011], although positive responses to $eC_a$ in terms of productivity were observed at all six sites. Interestingly, Norby and Zak [-@Norby2011] found a negative correlation between pre-treatment *L* and the responsiveness to $eC_a$, leading to the prediction that stands with low *L* will show a larger response to $eC_a$. Because diminished returns in increasing *L* in response to $eC_a$ when LAI is larger [@Norby2005].

\
In ecosystems where water availability limits productivity, we could expect $eC_a$ to further affect *L* [@Donohue2009], since $eC_a$ reduces stomatal conductance, and hence water use per unit of leaf area [@Warren2010] leading to increased soil water availability. These ‘water savings’ can be expected to lead to an increase in *L* in water-limited ecosystems under the idea that leaf area is closely coupled to water-balance. **Explain here or just before that we expect this because of the tight relationship between LAI and rainfall across gradients.** In support of this prediction, Donohue et al. [-@Donohue2013] attributed the observed global increase in *L* in arid and semi-arid areas over the last 30 years to rising $C_a$. Based on the same water-savings principle, we predict that *L* should also increase under $eC_a$ in water-limited forests.

@@

\
We estimated *L* from near-continuous measurements of canopy transmittance in a Eucalyptus-dominated woodland exposed to $eC_a$, the EucFACE experiment. The woodland stand at the EucFACE is frequently water-limited, allowing us to quantify the dynamics of *L*, and its components (litter fall and production) in response to water availability. We hypothesized that $eC_a$ would result in an increase in *L*, particularly during periods of water limitation, as water savings were expected to lead to longer retention of foliage under elevated $CO_2$. In addition, we expected that the low *L* that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allowed a larger response to $eC_a$ [@Norby2011]. Our second objective was to quantify the temporal dynamics of *L*, in particular in relation to fluctuating water availability.

@@

## Site description and experimental design
The EucFACE experiment is located on an ancient alluvial ﬂoodplain about 5 km from the Hawkesbury River in western Sydney (Australia) at 30 m a.s.l. elevation (33°37'S, 150°44'E). It is located within a remnant patch of native Cumberland Plain woodland (Tozer, 2003). The vegetation, maintained in this location since the 1880’s, retains some old-growth trees and some re-growth vegetation though it has been undisturbed for ca. 70 years. The site is characterized by a humid temperate-subtropical transitional climate with a mean annual temperature of 17 °C and mean annual precipitation of 800 mm (Bureau of Meteorology, station 067105 in Richmond, NSW Australia; http://www.bom.gov.au).

\
The soil at the site is a loamy sand in the Clarendon Formation (Barton et al., 2010; Isbell, 2002), a slightly-acid, low-organic C, well-drained soil. The soil surface is a loamy sand with > 75% sand content, and from ca. 50 to 300 cm depth, soils are a sandy clay loam with > 30% silt and clay. The soil is nutrient poor, and a strong limitation of P on tree growth at the site has recently been demonstrated [@Crous2015]. Groundwater depth is 12.6 – 12.9 m below the soil surface. The woodland ranges in density from 600-1000 trees $ha^{-1}$. The vegetation consists of an open woodland (basal area before treatment, BA = `r mean(ba$BA)`, SE = `r sd(ba$BA)/sqrt(6)` m$^2$ ha$^{-1}$, n = 6 plots) with a single major canopy-forming tree (*Eucalyptus tereticornis* Sm.) with a minor component of *E. amplifolia* Naudin. The understory stratum is composed mainly of grass species dominated by *Microlaena stipoides* (Labill.) R.Br., with lower densities of forbs of *Commelina cyanea* R.Br., *Lobelia purpurascens* R.Br., *Dichondra repens* J.R.Forst. & G.Forst and occasional shrubs of *Breynia oblongifolia* (Mull.Arg.) Mull.Arg. and *Bursaria spinosa* Cav.

\
Within the EucFACE site, there are six 25-m diameter circular plots (designated hereafter as ‘rings’). Each ring comprises 32 vent pipes in a cylindrical frame extending above the tree canopy to 28 m high. Each vertical pipe is subtended by a control valve and is connected to a circular plenum [@Hendrey1999] through which a computer-controlled system modulates release of diluted $CO_2$ into the vegetation according to a proportional-integral-differential type of algorithm [@Lewin2009]. The vegetation within three randomly selected rings was exposed to a $CO_2$ mole fraction ($C_a$) higher than ambient starting in September 2012, whereas the other three rings were identical to the $CO_2$ treatment rings in infrastructure and instrumentation but received ambient $C_a$, and are used as control plots. The target diurnal $C_a$ for the rings exposed to $eC_a$ was 150 μmol $CO_2$ mol$^{-1}$ above ambient $C_a$ in full operation mode. However, $C_a$ was increased gradually for six months at a rate of ca. 30 μmol $CO_2$ mol$^{-1}$ month$^{-1}$ until full operation mode was reached (Fig. 1a).

@@

## Leaf litter production
In each ring, 8 litter cloth baskets with an area of 0.1979 $m^2$ were installed at 0.5m above the forest floor. Litter was collected approximately monthly, and leaves separated from other components such as twigs, bark, seeds, insect frass, etc. Litter production was converted to units of *L* using an estimate of specific leaf area (SLA) determined from 329 leaves collected from the tree canopies from May 2012 to May 2014 (`r get_sla()` cm$^2$ g$^{-1}$, SE=2.14). We did not find a significant effect of $C_a$ treatment on SLA (Ellsworth, unpublished data; Ben Moore, unpublished data).

@@

where $\tau_d$ is diffuse canopy transmittance, Qabove is above canopy PAR (µmol m$^{-2}s^{-1}$), and Q$_{below}$ is the average PAR across the three sub-canopy PAR sensors in each FACE ring. We then calculated daily averages of $\tau_d$, excluding days where less than 4 time steps were available, and days where the standard deviation among the three PAR sensors in a ring was  > 0.03 (in units of $\tau_d$).

@@

where $\theta$ is the solar zenith angle, G($\theta$) is the foliage projection function in the direction of the beam [-@Campbell2000]. Because we assumed a spherical leaf angle distribution, G(θ) = 0.5 [@Stenberg2006]. We solved numerically for *L* in Eq. 3, using a smooth integration over $\theta$, yielding $L_\tau$.

@@

## Data analysis
We used generalized additive models (gam) for all analyses of *L* and $\tau_d$ timeseries [@Wood2006], using the *mgcv* package in R [@RCT2015]. For all time-series, we used a base of `r .gambase` when fitting the gams, which resulted in biologically realistic smoothed *L* dynamics. In all fits, we used the FACE ring as a random effect, except when we analysed average *L* across all rings (Fig. 7). We estimated the derivative of the fitted gam, including confidence intervals for the derivative, following Simpson [@Simpson2014]. For comparison between *L* from photos and PAR transmittance, we used simple linear regression. To test for effects of $C_a$ treatment on $L_\tau$ and litter production, we used linear-mixed effects models with FACE ring as the random effect, and $C_a$ treatment and date as a factor.

\
To assess effects of $eC_a$ on $L_\tau$, we accounted for pre-treatment differences in canopy structure, as follows. The average $L_\tau$ over the entire period was well correlated with basal area (BA) measured prior to the onset of the $eC_a$ treatment across the FACE rings (Fig. 3). Because the EucFACE woodland is well established, the differences in BA are not due to $C_a$ treatment effects, but instead represent natural variation within the woodland. We then calculated the deviation of actual $L_\tau$ from the fitted relationship with BA shown in Fig. 3, to yield a ‘*L* anomaly’. This way, we effectively used BA as a covariate in the analysis to account for pre-treatment variation.

@@

\
We found no differences in the *L* anomaly between the $C_a$ treatments, as shown by consistent overlap of the the 95% confidence intervals of the generalized additive model  (thus, p > 0.05 at any timepoint, Fig. 3). A brief exception was the period where one of the rings was affected by an insect outbreak (see above). 
A linear mixed-effects model with $C_a$ and time as fixed effects, and FACE ring as a random effect also showed no significant effect of $C_a$ on the *L* anomaly (p = `r LanomAOV[[3]][1]`). What difference between $C_a$ treatments could we have reasonably detected? The average half-width of the confidence interval on the *L* anomaly (Fig. 4) was `r LanomCIamb` $m^2$ $m^{-2}$ for $aC_a$, and `r LanomCIele` $m^2$ $m^{-2}$ for the $eC_a$ treatment. This means that, provided the within-treatment variance stays the same, we could detect a difference of `r LanomCIamb + LanomCIele` $m^2$ $m^{-2}$ (the sum of the half-width CI’s), or about a change of `r Ldetect`% over the mean *L* of the EucFACE (`r mean(ba$LAI.mean)` $m^2$ $m^{-2}$).

@@

\
We estimated gross leaf production from measured litter production ($P_L$) and temporal dynamics of $L_\tau$. Briefly, the change in $L_\tau$ was estimated for each litter collection period (again, from fitted generalized additive models to allow interpolation from a smoothed timeseries), and added to $P_L$ to arrive at an estimate of the leaf production. The temporal dynamics of litter and leaf production are shown in Fig. 5, and demonstrate the irregular temporal pattern in leaf production and $P_L$. We found no significant difference in either litter or leaf production between $C_a$ treatments (p > 0.2 for both, assessed with linear-mixed effects models).

\
Next, we related $P_L$ to average $L_\tau$ (during the litter collection period) across the six FACE rings (Fig. 6a). If leaf lifespan is constant, $P_L$ would be linearly related to $L_\tau$. Instead, we found that in $eC_a$, $P_L$ was lower at a given $L_\tau$ as compared to $aC_a$. We estimated mean leaf lifespan from Fig. 6a as the ratio of $L_\tau$ to annual $P_L$. We found a significant (p = `r LLtest$p.value`) increase in LL with elevated $C_a$ (Fig. 6b), with LL = `r LLtest$estimate1`yr in $aC_a$ and `r LLtest$estimate2`yr in $eC_a$.

@@

## Response of *L* to elevated [$CO_2$]
We report on `r nummon` months of continuous leaf area index (*L*) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated $CO_2$ ($eC_a$) and 3 months of ramp-up in $C_a$. We had predicted that *L* would increase in response to $eC_a$ as a result of increased quantum yield for photosynthesis, enhanced $CO_2$ assimilation for the entire canopy, and increased soil water availability. We also expected that this eucalypt woodland, with its inherently low *L* [@Whitehead2004], would be even more responsive to $eC_a$ than higher-*L* and more closed-canopy stands [@Norby2011]. Counter to our expectations but similar to Norby et al. [-@Norby2003], we found no significant effect of elevated $C_a$ on *L*. It could be argued that the change in *L* in response to $eC_a$ was too small to be detected. Natural variation in *L* among plots in a native woodland is indeed a challenge to manipulative studies with low replication, such as FACE experiments. To increase sensitivity and statistical robustness, we used a novel method to account for pre-treatment differences by using stand basal area as a covariate. Using this method, we estimated that we can detect a difference of only a `r format(Ldetect,digits=1)`% in *L*. Based on this small detectable difference, we conclude that the lack of a significant $C_a$ effect is not due to a lack of power, but a true lack of response to elevated $C_a$.

\
This lack of a response to elevated $C_a$ counters our expectations and those in previous publications [@Norby2003; @McCarthy2007], for a number of reasons. First, in a review of forest FACE sites, Norby and Zak [-@Norby2011] showed that low *L* stands were more responsive to $eC_a$. They argued that high *L* canopies would not benefit from further increases in *L*, because it would lead to only very small increases in total absorbed radiation. The mean *L* at the EucFACE was ca. `r mean(ba$LAI.mean)` $m^2$ $m^{-2}$, which is low compared to a global compendium of forest *L* [@Iio2013], and the lowest *L* of any forest FACE experiment [@Norby2011], but in line with a range of values for native *Eucalyptus* forests and woodlands reported by Whitehead and Beadle [-@Whitehead2004] and Iio et al. [-@Iio2013]. Based on this *L* estimate and the relationship across FACE sites reported by Norby and Zak [-@Norby2011], we could expect about a 30% increase in *L* under $eC_a$. It is clearly too early to draw general conclusions across FACE sites in terms of *L* response, moreover because water and nutrient likely play a central role in productivity and *L* dynamics at the EucFACE. Indeed, experimental evidence clearly demonstrates strong limitations of *L* by nutrients and water in Eucalyptus stands [@White2010].

\
A possible explanation for the lack of a *L* response to $eC_a$ is that leaf growth may not be directly related to increased carbon availability as a result of $eC_a$. In other words, stimulation of net leaf photosynthesis by $eC_a$ (a common response, also observed at the EucFACE, Gimeno et al. in review) may not lead to increased net leaf growth, due to other constraints such as water or nutrient limitations [@Crous2015]. Net leaf growth is the net sum of allocation of photosynthates to leaf growth, and turnover of the leaf pool. Either process may itself be affected by $eC_a$ [@DeKauwe2014]. We found a significant effect of $eC_a$ on leaf lifespan, with leaves living approximately `r round(abs(LLtest$estimate*12),1)` months longer in elevated $C_a$ (Fig. 6). Because we found no difference in *L* with $C_a$ treatment, this suggests a compensating difference in the fractional allocation of net primary production to leaf growth. To further understand the response of biomass allocation to $eC_a$ at the EucFACE, all components of the carbon balance must be carefully quantified, a task underway but not yet complete. For example, Drake et al. (in review) reported increased soil respiration under $eC_a$, suggesting that at least some of the stimulation of photosynthesis was allocated to either root growth, respiration, or root exudation.

@@

>>>>

---
title: Canopy leaf area of an evergreen Eucalyptus woodland does not respond to elevated
  atmospheric [CO~2~] but tracks water availability
author: ''
date: ''
output:
  word_document:
    reference_docx: manuscriptstyle.docx
  pdf_document: default
  html_document:
    number_sections: yes
csl: global-change-biology.csl
bibliography: references.bib
---

@@

library(broom)
library(lme4)
library(lmerTest)
library(car)
library(reporttools)

# shortcut to formatPval
pval <- function(...)formatPval(..., includeEquality=TRUE)

@@

# Some definitions of constants
.gambase <- 18
.maxSD <- 0.03

@@

# LAI ~ BA
balm <- lm(LAI.mean ~ BA, data=ba)

@@

# Canopy photo - Ltau comparison
lmcanphot <- lm(LAI.PAR.mean ~ LAI, data=flatcan_byring)

# With optimized k
flatcan_byring2 <- agg_flatcan(flatcan, by="Ring", k=kopt)
flatcan_byring2 <- add_PARLAI_to_flatcan(facegap_cloudy_byring,flatcan_byring2)
lmcanphot2 <- lm(LAI.PAR.mean ~ LAI, data=flatcan_byring2)

@@

\
Remko A. Duursma^1^, Teresa E. Gimeno^1,2^, Matthias M. Boer^1^, Kristine Y. Crous^1^, Mark G. Tjoelker^1^, David S. Ellsworth^1^.

\
^1^ Hawkesbury Institute for the Environment, University of Western Sydney, Locked Bag 1797, Penrith, NSW, Australia

\
^2^ INRA, UMR 1391 ISPA, 33140 Villenave d'Ornon, France

@@

Leaf area, quantified by the leaf area index (*L*), is a crucial driver of forest productivity, water use, and energy balance. Because *L* responds to environmental drivers, it can represent an important feedback to climate change, but its responses to rising CO~2~ and water availability of unmanaged forests have been poorly quantified. We studied canopy leaf area dynamics for `r nummon` months in a native evergreen *Eucalyptus* woodland exposed to Free Air CO~2~ Enrichment (the EucFACE experiment), in a temperate (hot summer) climate where water limitation is common. We hypothesized that, because of expected higher productivity and increased water-use efficiency, *L* should increase with elevated [CO~2~]. We estimated *L* from diffuse canopy transmittance, and measured monthly leaf litter production. Contrary to expectation, *L* did not respond to elevated [CO~2~] in this timeframe. We found that *L* varied between `r rangeLAI[1]` and `r rangeLAI[2]`, but did not show a clear annual pattern. Instead, dynamics of *L* showed a quick increase after heavy rainfall, and a steady during periods of low rainfall. Leaf litter production was correlated to changes in *L*, both during periods of decreasing *L* (when no leaf growth occurred), and during periods of increasing *L* (active shedding of old foliage when new leaf growth occurred). Leaf lifespan, estimated from mean *L* and total annual litter production, was significantly increased under elevated [CO~2~] (p `r pval(LLtest$p.value)`). We discuss these results in the context of modelling phenology of evergreen woodlands in response to environmental drivers.

@@

Leaf area index (*L*) is an important ecosystem attribute underlying productivity of vegetation, [@Reich2012], and a key variable in land surface models [@Kala2013; @Walker2014]. *L* is a dynamic canopy property, as it is generally responsive to variation in water and nutrient availability, as demonstrated in experiments [@Herbert1995; @Hubbard2004; @White2010] and observational studies across rainfall gradients [@Ellis2008; @Iio2013; @Jin2013]. Ecosystem models generally predict that *L* increases in response to elevated atmospheric [CO~2~] (eC~a~) due to increased net carbon uptake [@DeKauwe2014; @Hatton1992; @Kergoat2002; @Zhu2012]. Also expected because increased quantum yield of photosynthesis in eC~a~ would allow leaf retention into deeper shade [@Norby1999], but little empirical evidence is available on the response of *L* to eC~a~ in mature forests [@Norby2011]. Quantifying the magnitude and direction of the eC~a~ effect on *L* is crucial, as changes in *L* associated with climate change drivers such as increases in C~a~ may provide an important feedback to climate change, and modify direct physiological responses to C~a~ [@Tor-ngern2015; @Murray2012].

\
Direct effects of eC~a~ on vegetation processes include an enhancement of photosynthesis and productivity, a decrease in stomatal conductance and leaf-level water use and an increase in the water-use efficiency [@Ainsworth2005; @Ellsworth2012]. Although numerous studies have been conducted on the response of physiology and plant growth to eC~a~, very few of these inform the response of steady-state *L* to eC~a~, because it requires a sufficiently mature forest to be exposed to free air CO~2~ enrichment (FACE) [@Norby2011]. Increased carbon availability through enhanced photosynthesis under eC~a~ does not necessarily lead to increases in *L*. Out of the six forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch) and needle-leaved evergreen pine stands showed increased *L* in response to eC~a~ *L* [@Norby2011], although positive responses to eC~a~ in terms of productivity were observed at all six sites. Interestingly, Norby and Zak [-@Norby2011] found a negative correlation between pre-treatment *L* and the responsiveness to eC~a~, leading to the prediction that stands with low *L* will show a larger response to eC~a~. Because diminished returns in increasing *L* in response to eC~a~ when LAI is larger [@Norby2005].

\
In ecosystems where water availability limits productivity, we could expect eC~a~ to further affect *L* [@Donohue2009], since eC~a~ reduces stomatal conductance, and hence water use per unit of leaf area [@Warren2010] leading to increased soil water availability. These ‘water savings’ can be expected to lead to an increase in *L* in water-limited ecosystems under the idea that leaf area is closely coupled to water-balance. **Explain here or just before that we expect this because of the tight relationship between LAI and rainfall across gradients.** In support of this prediction, Donohue et al. [-@Donohue2013] attributed the observed global increase in *L* in arid and semi-arid areas over the last 30 years to rising C~a~. Based on the same water-savings principle, we predict that *L* should also increase under eC~a~ in water-limited forests.

@@

\
We estimated *L* from near-continuous measurements of canopy transmittance in a Eucalyptus-dominated woodland exposed to eC~a~, the EucFACE experiment. The woodland stand at the EucFACE is frequently water-limited, allowing us to quantify the dynamics of *L*, and its components (litter fall and production) in response to water availability. We hypothesized that eC~a~ would result in an increase in *L*, particularly during periods of water limitation, as water savings were expected to lead to longer retention of foliage under elevated CO~2~. In addition, we expected that the low *L* that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allowed a larger response to eC~a~ [@Norby2011]. Our second objective was to quantify the temporal dynamics of *L*, in particular in relation to fluctuating water availability.

@@

## Site description and experimental design
The EucFACE experiment is located on an ancient alluvial floodplain about 5 km from the Hawkesbury River in western Sydney (Australia) at 30 m a.s.l. elevation (33°37'S, 150°44'E). It is located within a remnant patch of native Cumberland Plain woodland (Tozer, 2003). The vegetation, maintained in this location since the 1880’s, retains some old-growth trees and some re-growth vegetation though it has been undisturbed for ca. 70 years. The site is characterized by a humid temperate-subtropical transitional climate with a mean annual temperature of 17 °C and mean annual precipitation of 800 mm (Bureau of Meteorology, station 067105 in Richmond, NSW Australia; http://www.bom.gov.au).

\
The soil at the site is a loamy sand in the Clarendon Formation (Barton et al., 2010; Isbell, 2002), a slightly-acid, low-organic C, well-drained soil. The soil surface is a loamy sand with > 75% sand content, and from ca. 50 to 300 cm depth, soils are a sandy clay loam with > 30% silt and clay. The soil is nutrient poor, and a strong limitation of P on tree growth at the site has recently been demonstrated [@Crous2015]. Groundwater depth is 12.6 – 12.9 m below the soil surface. The woodland ranges in density from 600-1000 trees ha^-1^. The vegetation consists of an open woodland (basal area before treatment, BA = `r mean(ba$BA)`, SE = `r sd(ba$BA)/sqrt(6)` m^2^ha^-1^, n = 6 plots) with a single major canopy-forming tree (*Eucalyptus tereticornis* Sm.) with a minor component of *E. amplifolia* Naudin. The understory stratum is composed mainly of grass species dominated by *Microlaena stipoides* (Labill.) R.Br., with lower densities of forbs of *Commelina cyanea* R.Br., *Lobelia purpurascens* R.Br., *Dichondra repens* J.R.Forst. & G.Forst and occasional shrubs of *Breynia oblongifolia* (Mull.Arg.) Mull.Arg. and *Bursaria spinosa* Cav.

\
Within the EucFACE site, there are six 25-m diameter circular plots (designated hereafter as ‘rings’). Each ring comprises 32 vent pipes in a cylindrical frame extending above the tree canopy to 28 m high. Each vertical pipe is subtended by a control valve and is connected to a circular plenum [@Hendrey1999] through which a computer-controlled system modulates release of diluted CO~2~ into the vegetation according to a proportional-integral-differential type of algorithm [@Lewin2009]. The vegetation within three randomly selected rings was exposed to a CO~2~ mole fraction (C~a~) higher than ambient starting in September 2012, whereas the other three rings were identical to the CO~2~ treatment rings in infrastructure and instrumentation but received ambient C~a~, and are used as control plots. The target diurnal C~a~ for the rings exposed to eC~a~ was 150 $\mu$mol CO~2~ mol^-1^ above ambient C~a~ in full operation mode. However, C~a~ was increased gradually for six months at a rate of ca. 30 $\mu$mol CO~2~ mol^-1^ month^-1^ until full operation mode was reached (Fig. 1a).

@@

## Leaf litter production
In each ring, 8 litter cloth baskets with an area of 0.1979 m^2^ were installed at 0.5m above the forest floor. Litter was collected approximately monthly, and leaves separated from other components such as twigs, bark, seeds, insect frass, etc. Litter production was converted to units of *L* using an estimate of specific leaf area (SLA) determined from 329 leaves collected from the tree canopies from May 2012 to May 2014 (`r get_sla()` cm^2^ g^-1^, SE=2.14). We did not find a significant effect of C~a~ treatment on SLA (Ellsworth, unpublished data; Ben Moore, unpublished data).

@@

where $\tau_d$ is diffuse canopy transmittance, Q~above~ is above canopy PAR (µmol m^-2^s^-1^), and Q~below~ is the average PAR across the three sub-canopy PAR sensors in each FACE ring. We then calculated daily averages of $\tau_d$, excluding days where less than 4 time steps were available, and days where the standard deviation among the three PAR sensors in a ring was > `r .maxSD` (in units of $\tau_d$).

@@

where $\theta$ is the solar zenith angle, G($\theta$) is the foliage projection function in the direction of the beam [-@Campbell2000]. Because we assumed a spherical leaf angle distribution, G($\theta$) = 0.5 [@Stenberg2006]. We solved numerically for *L* in Eq. 3, using a smooth integration over $\theta$, yielding $L_\tau$.

@@

## Data analysis
We used generalized additive models (gam) for all analyses of *L* and $\tau_d$ timeseries [@Wood2006], using the *mgcv* package in R [@RCT2015]. For all time-series, we used a base of `r .gambase` when fitting the gams, which resulted in biologically realistic smoothed *L* dynamics. In all fits, we used the FACE ring as a random effect, except when we analysed average *L* across all rings (Fig. 7). We estimated the derivative of the fitted gam, including confidence intervals for the derivative, following Simpson [@Simpson2014]. For comparison between *L* from photos and PAR transmittance, we used simple linear regression. To test for effects of C~a~ treatment on $L_\tau$ and litter production, we used linear-mixed effects models with FACE ring as the random effect, and C~a~ treatment and date as a factor.

\
To assess effects of eC~a~ on $L_\tau$, we accounted for pre-treatment differences in canopy structure, as follows. The average $L_\tau$ over the entire period was well correlated with basal area (BA) measured prior to the onset of the eC~a~ treatment across the FACE rings (Fig. 3). Because the EucFACE woodland is well established, the differences in BA are not due to C~a~ treatment effects, but instead represent natural variation within the woodland. We then calculated the deviation of actual $L_\tau$ from the fitted relationship with BA shown in Fig. 3, to yield a ‘*L* anomaly’. This way, we effectively used BA as a covariate in the analysis to account for pre-treatment variation.

@@

\
We found no differences in the *L* anomaly between the C~a~ treatments, as shown by consistent overlap of the the 95% confidence intervals of the generalized additive model  (thus, p > 0.05 at any timepoint, Fig. 3). A brief exception was the period where one of the rings was affected by an insect outbreak (see above). 
A linear mixed-effects model with C~a~ and time as fixed effects, and FACE ring as a random effect also showed no significant effect of C~a~ on the *L* anomaly (p `r pval(LanomAOV[[3]][1])`). What difference between C~a~ treatments could we have reasonably detected? The average half-width of the confidence interval on the *L* anomaly (Fig. 4) was `r LanomCIamb` m^2^ m^-2^ for aC~a~, and `r LanomCIele` m^2^m^-2^ for the eC~a~ treatment. This means that, provided the within-treatment variance stays the same, we could detect a difference of `r LanomCIamb + LanomCIele` $m^2$ $m^{-2}$ (the sum of the half-width CI’s), or about a change of `r Ldetect`% over the mean *L* of the EucFACE (`r mean(ba$LAI.mean)` m^2^m^-2^).

@@

\
We estimated gross leaf production from measured litter production ($P_L$) and temporal dynamics of $L_\tau$. Briefly, the change in $L_\tau$ was estimated for each litter collection period (again, from fitted generalized additive models to allow interpolation from a smoothed timeseries), and added to $P_L$ to arrive at an estimate of the leaf production. The temporal dynamics of litter and leaf production are shown in Fig. 5, and demonstrate the irregular temporal pattern in leaf production and $P_L$. We found no significant difference in either litter or leaf production between C~a~ treatments (p > 0.2 for both, assessed with linear-mixed effects models).

\
Next, we related $P_L$ to average $L_\tau$ (during the litter collection period) across the six FACE rings (Fig. 6a). If leaf lifespan is constant, $P_L$ would be linearly related to $L_\tau$. Instead, we found that in eC~a~, $P_L$ was lower at a given $L_\tau$ as compared to aC~a~. We estimated mean leaf lifespan from Fig. 6a as the ratio of $L_\tau$ to annual $P_L$. We found a significant (p `r pval(LLtest$p.value)`) increase in LL with elevated C~a~ (Fig. 6b), with LL = `r LLtest$estimate1`yr in aC~a~ and `r LLtest$estimate2`yr in eC~a~.

@@

## Response of *L* to elevated [CO~2~]
We report on `r nummon` months of continuous leaf area index (*L*) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated CO~2~ (eC~a~) and 3 months of ramp-up in C~a~. We had predicted that *L* would increase in response to eC~a~ as a result of increased quantum yield for photosynthesis, enhanced CO~2~ assimilation for the entire canopy, and increased soil water availability. We also expected that this eucalypt woodland, with its inherently low *L* [@Whitehead2004], would be even more responsive to eC~a~ than higher-*L* and more closed-canopy stands [@Norby2011]. Counter to our expectations but similar to Norby et al. [-@Norby2003], we found no significant effect of elevated C~a~ on *L*. It could be argued that the change in *L* in response to eC~a~ was too small to be detected. Natural variation in *L* among plots in a native woodland is indeed a challenge to manipulative studies with low replication, such as FACE experiments. To increase sensitivity and statistical robustness, we used a novel method to account for pre-treatment differences by using stand basal area as a covariate. Using this method, we estimated that we can detect a difference of only a `r format(Ldetect,digits=1)`% in *L*. Based on this small detectable difference, we conclude that the lack of a significant C~a~ effect is not due to a lack of power, but a true lack of response to elevated C~a~.

\
This lack of a response to elevated C~a~ counters our expectations and those in previous publications [@Norby2003; @McCarthy2007], for a number of reasons. First, in a review of forest FACE sites, Norby and Zak [-@Norby2011] showed that low *L* stands were more responsive to eC~a~. They argued that high *L* canopies would not benefit from further increases in *L*, because it would lead to only very small increases in total absorbed radiation. The mean *L* at the EucFACE was ca. `r mean(ba$LAI.mean)` m^2^m^-2^, which is low compared to a global compendium of forest *L* [@Iio2013], and the lowest *L* of any forest FACE experiment [@Norby2011], but in line with a range of values for native *Eucalyptus* forests and woodlands reported by Whitehead and Beadle [-@Whitehead2004] and Iio et al. [-@Iio2013]. Based on this *L* estimate and the relationship across FACE sites reported by Norby and Zak [-@Norby2011], we could expect about a 30% increase in *L* under eC~a~. It is clearly too early to draw general conclusions across FACE sites in terms of *L* response, moreover because water and nutrient likely play a central role in productivity and *L* dynamics at the EucFACE. Indeed, experimental evidence clearly demonstrates strong limitations of *L* by nutrients and water in Eucalyptus stands [@White2010].

\
A possible explanation for the lack of a *L* response to eC~a~ is that leaf growth may not be directly related to increased carbon availability as a result of eC~a~. In other words, stimulation of net leaf photosynthesis by eC~a~ (a common response, also observed at the EucFACE, Gimeno et al. in review) may not lead to increased net leaf growth, due to other constraints such as water or nutrient limitations [@Crous2015]. Net leaf growth is the net sum of allocation of photosynthates to leaf growth, and turnover of the leaf pool. Either process may itself be affected by eC~a~ [@DeKauwe2014]. We found a significant effect of eC~a~ on leaf lifespan, with leaves living approximately `r round(abs(LLtest$estimate*12),1)` months longer in elevated C~a~ (Fig. 6). Because we found no difference in *L* with C~a~ treatment, this suggests a compensating difference in the fractional allocation of net primary production to leaf growth. To further understand the response of biomass allocation to eC~a~ at the EucFACE, all components of the carbon balance must be carefully quantified, a task underway but not yet complete. For example, Drake et al. (in review) reported increased soil respiration under eC~a~, suggesting that at least some of the stimulation of photosynthesis was allocated to either root growth, respiration, or root exudation.

@@

# Figure captions {.unnumbered}
